Characteristic polynomials of p-adic matrices

### 2017-02-04 charpoly/charpoly.tex

\todo{Explain here why computing characteric polynomial is so important}

\todo{Explain also why taking care of precision is important}

\todo{Example of $M$ and $M + 1$}

\noindent
{\bf Previous contributions.}
\todo{Many \emph{ad-hoc} for Kedlaya-like counting point algorithms.
Our previous paper. What else?}

\noindent
{\bf The contribution of this paper.}

\noindent
{\bf Organization of the article.}

\noindent
{\bf Notations.}
Throughout the paper, $K$ will refer to a complete, discrete valuation 
field, $\val : K \twoheadrightarrow \Z \cup \{+\infty\}$ to its 
valuation, $\OK$ its ring of integers and $\pi$ a uniformizer.

@@

Let us consider more precisely the case where $M$ is given at 
jagged precision: the $(i,j)$ entry of $M$ is given at precision 
$O(\pi^{N_{i,j}})$ for some given integer $N_{i,j}$. 
Lemma 3.4 of \cite{caruso-roe-vaccon:14a} then teaches us that
the optimal precision on the $k$-th coefficient of $\chi_M$ is 
$O(\pi^{N'_k})$ where $N'_k$ is given by the formula:
\begin{equation}
\label{eq:optjagged}
N'_k = \min_{1 \leq i, j\leq n} N_{j,i} + \val(\pi_k(C_{i,j}))
\end{equation}
where $C_{i,j}$ is the $(i,j)$ entry of the comatrix $\com(X{-}M)$.

@@

\begin{rem}
If $M \in M_n(O_K)$ is given at (high enough) \emph{flat} precision, 
then we can avoid the final base change step in the Hessenberg method.
Indeed, observe that, thanks to Lemma \ref{lem:comatrix_of_similar}, 
we can write:
$$\tr(\com (X{-}M) \cdot dM)=\tr(\com (X{-}H)\cdot P^{-1} dM P)$$
where $P$ lies in $\GL_n(\OK)$. Moreover, the latter condition implies
that $P^{-1} dM P$ runs over $M_n(\OK)$ when $P$ runs over $M_n(\OK)$.
As a consequence, the integer $N'_k$ given the optimal precision on the 
$k$-th coefficient of $M$ is also equal to
$N + \min_{1 \leq i, j\leq n} \val(\pi_k(C^H_{i,j}))$
where $C^H_{i,j}$ is the $(i,j)$ entry of the comatrix of $X{-}H$
where $H$ is the Hessenberg form of $M$.
\end{rem}

@@

\noindent
{\bf Numerical experiments.}
We have made numerical experiments in \textsc{SageMath}~\cite{sage}
in order to compare the optimal precision obtained with the methods
explained above with the actual precision obtained by the software.
For doing so, we picked a sample of $1000$ random matrices $M$ in 
$M_9(\Q_2)$ where all the entries at the same \emph{relative
precision}.
We recall that, in \textsc{SageMath}, random elements $x \in \Q_p$ are 
generated as follows. We fix an integer $N$ --- the so-called relative
precision --- and generate elements of $\Q_p$ of the shape
$$x = p^v \cdot \big(a + O\big(p^{N+v_p(a)}\big)\big)$$
where $v$ is a random integer generated according to the distribution:
$$\mathbb P [v = 0] = \frac 1 5 \quad ; \quad
\mathbb P [v = n] = \frac 2 {5\cdot |n| \cdot (|n|+1)} \text{ for }
n \geq 1$$
and $a$ is a random integer in the range $[0, p^N)$ with uniform

@@

\noindent
The table of Figure~\ref{fig:exp} summarizes the obtaines results. 
%
\begin{figure}
\hfill
{\renewcommand*{\arraystretch}{1.3}
\begin{tabular}{|c|r|r|r|}
\cline{2-4} 
\multicolumn{1}{l|}{} 
  & \multicolumn{3}{c|}{Average loss of accuracy} \\
\cline{2-4}
\multicolumn{1}{l|}{\null\hspace{6mm}\null} 
  & \makebox[1.5cm]{\hfill Optimal\hfill\null}
  & \makebox[1.5cm]{\hfill CR\hfill\null}
  & \makebox[1.5cm]{\hfill FP\hfill\null} \\
\hline
\rule{0pt}{2.7ex}%
$X^0$ & $3.17$ & $196\phantom{.00}$ & $189\phantom{.00}$ \vspace{-1.5ex} \\
{\scriptsize (det.)} 
& {\scriptsize dev: $1.76$} & {\scriptsize dev: $240$} & {\scriptsize dev: $226$} \\
\rule{0pt}{2.7ex}%
$X^1$ & $2.98$ & $161\phantom{.00}$ & $156\phantom{.00}$ \vspace{-1.5ex} \\
& {\scriptsize dev: $1.69$} & {\scriptsize dev: $204$} & {\scriptsize dev: $195$} \\
\rule{0pt}{2.7ex}%
$X^2$ & $2.75$ & $129\phantom{.00}$ & $126\phantom{.00}$ \vspace{-1.5ex} \\
& {\scriptsize dev: $1.57$} & {\scriptsize dev: $164$} & {\scriptsize dev: $164$} \\
\rule{0pt}{2.7ex}%
$X^3$ & $2.74$ & $108\phantom{.00}$ & $105\phantom{.00}$ \vspace{-1.5ex} \\
& {\scriptsize dev: $1.73$} & {\scriptsize dev: $144$} & {\scriptsize dev: $143$} \\
\rule{0pt}{2.7ex}%
$X^4$ & $2.57$ &  $63.2\phantom{0}$ &  $60.6\phantom{0}$ \vspace{-1.5ex} \\
& {\scriptsize dev: $1.70$} & {\scriptsize dev: $85.9$} & {\scriptsize dev: $85.8$} \\
\rule{0pt}{2.7ex}%
$X^5$ & $2.29$ &  $51.6\phantom{0}$ &  $49.7\phantom{0}$ \vspace{-1.5ex} \\
& {\scriptsize dev: $1.66$} & {\scriptsize dev: $75.3$} & {\scriptsize dev: $74.9$} \\
\rule{0pt}{2.7ex}%
$X^6$ & $2.07$ &   $9.04$           &   $8.59$ \vspace{-1.5ex} \\
& {\scriptsize dev: $1.70$} & {\scriptsize dev: $26.9$} & {\scriptsize dev: $26.4$} \\
\rule{0pt}{2.7ex}%
$X^7$ & $1.64$ &   $5.70$           &   $5.38$ \vspace{-1.5ex} \\
& {\scriptsize dev: $1.65$} & {\scriptsize dev: $15.3$} & {\scriptsize dev: $14.7$} \\
\rule{0pt}{2.7ex}%
$X^8$ & $0.99$ &   $0.99$           &   $0.99$ \vspace{-1.5ex} \\
{\scriptsize (trace)} 
& {\scriptsize dev: $1.37$} & {\scriptsize dev: $1.37$} & {\scriptsize dev: $1.37$} \\
\hline
\end{tabular}}
\hfill\null

@@

We observe that the loss of relative accuracy stays under control in the 
``Optimal'' column whereas it has a very erratic behavior --- very large 
values and very large deviation as well --- in the two other columns. 
These experiments then strongly support the interest of applying the 
methods developed in this paper.

@@

Let $M \in M_n(K)$ and $\lambda \in K$ be a \emph{simple} eigenvalue 
of $M$. We are interesting in quantifying the optimal precision on 
$\lambda$ when $M$ is given with some uncertainty.

For doing so, we fix an approximation $M_\app \in M_n(K)$ of $M$ and 
suppose that the uncertainty of $M$ is ``jagged'' in the sense that
each entry of $M$ is given at some precision $O(\pi^{N_{i,j}})$.
Let $\lambda_\app$ be the relevant eigenvalue of $M_\app$. We remark 
that it is possible to follow the eigenvalue $\lambda_\app$ on a small 
neighborhood $\calU$ of $M$. Precisely, there exists a unique continuous 
function $f : \calU \to K$ such that:
\begin{enumerate}[$\bullet$]
\renewcommand{\itemsep}{0pt}
\item $f(M_\app) = \lambda_\app$, and
\item $f(M)$ is an eigenvalue of $M$ for all $M \in \calU$.
\end{enumerate}

@@

\begin{proof}
The first assertion follows from the implicit function Theorem.
Differentiating the relation $\chi_M(\lambda) = 0$, we get
$\chi'_M(\lambda) \cdot d \lambda + \tr(\com(X-M) \cdot dM)(\lambda) = 0$,
from what the Lemma follows.
\end{proof}

\noindent
Lemma 3.4 of \cite{caruso-roe-vaccon:14a} now implies that, if the
$N_{i,j}$'s are large enough and rather well balanced, the optimal
precision on the eigenvalue $\lambda$ is $O(p^{N'})$ with:
$$N' = \min_{1 \leq i, j\leq n} \big(N_{j,i} + \val(C_{i,j}(\lambda)) - 
\val(\chi'_M(\lambda))\big)$$
where $C_{i,j}$ denotes as above the $(i,j)$ entry of $\com(X{-}M)$.
Writing
$\com(X{-}M) = \alpha \cdot P V^t \cdot V Q^t \text{ mod } \chi_M$
as in Proposition~\ref{prop:shortcom}, we find:
\begin{align*}
N' & = \val(\alpha(\lambda)) - \val(\chi'_M(\lambda)) \\
& \hspace{2mm} + \min_{1 \leq i, j\leq n} \big(N_{j,i} + 
\val(P_i V(\lambda)^t) + \val(V(\lambda) Q_j^t)\big)
\end{align*}
where $P_i$ denotes the $i$-th row of $P$ and, similarly, $Q_j$
denotes the $j$-th row of $Q$. Note moreover that $V(\lambda)$ is
the row vector $(1, \lambda, \ldots, \lambda^{n-1})$.
By the discussion of \S \ref{}, the exact value of $M$ can be 
determined for a cost of $\softO(n^\omega)$ operations in $K$ and
$O(n^2)$ operations on integers.

When $M$ is given at flat precision, \emph{i.e.} the $N_{i,j}$'s are all 
equal to some $N$, the formula for $N'$ rewrites:
\begin{align*}
N' & = N + \val(\alpha(\lambda)) - \val(\chi'_M(\lambda)) \\
& \hspace{2mm} + \min_{1 \leq i \leq n} \val(P_i V(\lambda)^t)
+ \min_{1 \leq j \leq n} \val(V(\lambda) Q_j^t)\big)
\end{align*}
and can therefore now be evaluated for a cost of $\softO(n^\omega)$
operations in $K$ and only $O(n)$ operations on integers.

>>>>

\todo{Explain here why computing characteristic polynomial is so important.}

The characteristic polynomial is a fundamental invariant of a matrix: its roots
give the eigenvalues, and the trace and determinant can be extracted from its coefficients.
In fact, the best known division-free algorithm for computing determinants
over arbitrary rings \cite{kartofen:92a} does so using the characteristic polynomial.
Over $p$-adic fields, computing the characteristic polynomial is a key ingredient in
algorithms for counting points of varieties over finite fields. \todo{list some point counting references.}
\todo{Application to characteristic polynomials of elements of p-adic field extensions?}

\todo{Explain also why taking care of precision is important.}

When computing with $p$-adic matrices, the lack of infinite memory implies
that the entries may only be approximated at some finite precision $O(p^m)$.  As a
consequence, in designing algorithms for such matrices one must analyze
not only the running time of the algorithm but also the accuracy of the result.

Let $M \in M_n(\Qp)$ be known at precision $O(p^m)$.
The simplest approach for computing the characteristic polynomial of $M$
is to compute $\det(X-M)$ either using recursive row expansion or various division free
algorithms \cites{\cite{seifullin:02a} \cite{kartofen:92a}}.  There are two issues
with these methods.  First, they are slower than alternatives that allow division,
requiring $O(n!)$, $O(n^4)$ and $\softO(n^{2 + \omega/2})$ operations.  Second,
while the lack of division implies that the result is accurate modulo $p^m$ as
long as $M \in M_n(\Zp)$, they still do not yield the optimal precision.

A faster approach over a field is to compute the Frobenius normal form of $M$,
which is achievable in running time $\softO(n^\omega)$ \cite{storjohann:01a}.
However, the fact that it uses division frequently leads to catastrophic losses of division.
In many examples, no precision remains at the end of the calculation.

Instead

\todo{Example of $M$ and $M + 1$.}

\subsection*{Previous contributions.} Many ad-hoc for Kedlaya-like point counting algorithms.  Our previous papers.  What else?

\subsection*{The contribution of this paper.} % Or \textbf

\subsection*{Organization of the article.}

\subsection*{Notation} Throughout the paper, $K$ will refer to a complete,
discrete valuation field, $\val : K \twoheadrightarrow \Z \cup \{+\infty\}$ to its valuation,
$\OK$ its ring of integers and $\pi$ a uniformizer. \todo{Soft-Oh}

@@

Let us consider more precisely the case where $M$ is given at 
jagged precision: the $(i,j)$ entry of $M$ is given at precision 
$O(\pi^{N_{i,j}})$ for some integers $N_{i,j}$. 
Lemma 3.4 of \cite{caruso-roe-vaccon:14a} then shows that
the optimal precision on the $k$-th coefficient of $\chi_M$ is 
$O(\pi^{N'_k})$ where $N'_k$ is given by the formula:
\begin{equation}
\label{eq:optjagged}
N'_k = \min_{1 \leq i, j\leq n} N_{j,i} + \val(\pi_k(C_{i,j})),
\end{equation}
where $C_{i,j}$ is the $(i,j)$ entry of the comatrix $\com(X{-}M)$.

@@

\begin{rem}
If $M \in M_n(O_K)$ is given at (high enough) \emph{flat} precision, 
then we can avoid the final base change step in the Hessenberg method.
Indeed, observe that, thanks to Lemma \ref{lem:comatrix_of_similar}, 
we can write:
$$\tr(\com (X{-}M) \cdot dM)=\tr(\com (X{-}H)\cdot P^{-1} dM P)$$
where $P$ lies in $\GL_n(\OK)$. Moreover, the latter condition implies
that $P^{-1} dM P$ runs over $M_n(\OK)$ when $P$ runs over $M_n(\OK)$.
As a consequence, the integer $N'_k$ giving the optimal precision on the 
$k$-th coefficient of $M$ is also equal to
$N + \min_{1 \leq i, j\leq n} \val(\pi_k(C^H_{i,j}))$
where $C^H_{i,j}$ is the $(i,j)$ entry of $\com(X{-}H)$,
where $H$ is the Hessenberg form of $M$.
\end{rem}

@@

\noindent
{\bf Numerical experiments.}
We have made numerical experiments in \textsc{SageMath}~\cite{sage}
in order to compare the optimal precision obtained with the methods
explained above with the actual precision obtained by the software.
For doing so, we picked a sample of $1000$ random matrices $M$ in 
$M_9(\Q_2)$ where all the entries at the same \emph{relative
precision}.
We recall that, in \textsc{SageMath}, random elements $x \in \Q_p$ are 
generated as follows. We fix an integer $N$ --- the so-called relative
precision --- and generate elements of $\Q_p$ of the shape
$$x = p^v \cdot \big(a + O\big(p^{N+v_p(a)}\big)\big)$$
where $v$ is a random integer generated according to the distribution:
$$\mathbb P [v = 0] = \frac 1 5 \quad ; \quad
\mathbb P [v = n] = \frac 2 {5\cdot |n| \cdot (|n|+1)} \text{ for }
|n| \geq 1$$
and $a$ is an integer in the range $[0, p^N)$, selected uniformly at random.

@@

\noindent
The table of Figure~\ref{fig:exp} summarizes the results obtained. 
%
\begin{figure}
\hfill
{\renewcommand*{\arraystretch}{1.3}
\begin{tabular}{|c|r|r|r|}
\cline{2-4} 
\multicolumn{1}{l|}{} 
  & \multicolumn{3}{c|}{Average loss of accuracy} \\
\cline{2-4}
\multicolumn{1}{l|}{\null\hspace{6mm}\null} 
  & \makebox[1.5cm]{\hfill Optimal\hfill\null}
  & \makebox[1.5cm]{\hfill CR\hfill\null}
  & \makebox[1.5cm]{\hfill FP\hfill\null} \\
\hline
\rule{0pt}{2.7ex}%
$X^0$ & $3.17$ & $196\phantom{.00}$ & $189\phantom{.00}$ \vspace{-1.5ex} \\
{\scriptsize (det.)} 
& {\scriptsize dev: $1.76$} & {\scriptsize dev: $240$} & {\scriptsize dev: $226$} \\
\rule{0pt}{2.7ex}%
$X^1$ & $2.98$ & $161\phantom{.00}$ & $156\phantom{.00}$ \vspace{-1.5ex} \\
& {\scriptsize dev: $1.69$} & {\scriptsize dev: $204$} & {\scriptsize dev: $195$} \\
\rule{0pt}{2.7ex}%
$X^2$ & $2.75$ & $129\phantom{.00}$ & $126\phantom{.00}$ \vspace{-1.5ex} \\
& {\scriptsize dev: $1.57$} & {\scriptsize dev: $164$} & {\scriptsize dev: $164$} \\
\rule{0pt}{2.7ex}%
$X^3$ & $2.74$ & $108\phantom{.00}$ & $105\phantom{.00}$ \vspace{-1.5ex} \\
& {\scriptsize dev: $1.73$} & {\scriptsize dev: $144$} & {\scriptsize dev: $143$} \\
\rule{0pt}{2.7ex}%
$X^4$ & $2.57$ &  $63.2\phantom{0}$ &  $60.6\phantom{0}$ \vspace{-1.5ex} \\
& {\scriptsize dev: $1.70$} & {\scriptsize dev: $85.9$} & {\scriptsize dev: $85.8$} \\
\rule{0pt}{2.7ex}%
$X^5$ & $2.29$ &  $51.6\phantom{0}$ &  $49.7\phantom{0}$ \vspace{-1.5ex} \\
& {\scriptsize dev: $1.66$} & {\scriptsize dev: $75.3$} & {\scriptsize dev: $74.9$} \\
\rule{0pt}{2.7ex}%
$X^6$ & $2.07$ &   $9.04$           &   $8.59$ \vspace{-1.5ex} \\
& {\scriptsize dev: $1.70$} & {\scriptsize dev: $26.9$} & {\scriptsize dev: $26.4$} \\
\rule{0pt}{2.7ex}%
$X^7$ & $1.64$ &   $5.70$           &   $5.38$ \vspace{-1.5ex} \\
& {\scriptsize dev: $1.65$} & {\scriptsize dev: $15.3$} & {\scriptsize dev: $14.7$} \\
\rule{0pt}{2.7ex}%
$X^8$ & $0.99$ &   $0.99$           &   $0.99$ \vspace{-1.5ex} \\
{\scriptsize (trace)} 
& {\scriptsize dev: $1.37$} & {\scriptsize dev: $1.37$} & {\scriptsize dev: $1.37$} \\
\hline
\end{tabular}}
\hfill\null

@@

We observe that the loss of relative accuracy stays under control in the 
``Optimal'' column whereas it has a very erratic behavior --- very large 
values and very large deviation as well --- in the two other columns. 
These experiments thus demonstrate the utility of the methods developed
in this paper.

@@

Let $M \in M_n(K)$ and $\lambda \in K$ be a \emph{simple}
\footnote{the corresponding generalized eigenspace has dimension $1$} eigenvalue 
of $M$. We are interesting in quantifying the optimal precision on 
$\lambda$ when $M$ is given with some uncertainty.

To do so, we fix an approximation $M_\app \in M_n(K)$ of $M$ and 
suppose that the uncertainty of $M$ is ``jagged'' in the sense that
each entry of $M$ is given at some precision $O(\pi^{N_{i,j}})$.
Let $\lambda_\app$ be the relevant eigenvalue of $M_\app$. We remark 
that it is possible to follow the eigenvalue $\lambda_\app$ on a small 
neighborhood $\calU$ of $M$. More precisely, there exists a unique continuous 
function $f : \calU \to K$ such that:
\begin{enumerate}[$\bullet$]
\renewcommand{\itemsep}{0pt}
\item $f(M_\app) = \lambda_\app$, and
\item $f(M')$ is an eigenvalue of $M'$ for all $M' \in \calU$.
\end{enumerate}

@@

\begin{proof}
The first assertion follows from the implicit function Theorem.
Differentiating the relation $\chi_M(\lambda) = 0$, we get
$\chi'_M(\lambda) \cdot d \lambda + \tr(\com(X-M) \cdot dM)(\lambda) = 0$,
from which the Lemma follows.
\end{proof}

\noindent
Lemma 3.4 of \cite{caruso-roe-vaccon:14a} now implies that, if the
$N_{i,j}$'s are large enough and sufficiently well balanced, the optimal
precision on the eigenvalue $\lambda$ is $O(p^{N'})$ with:
$$N' = \min_{1 \leq i, j\leq n} \big(N_{j,i} + \val(C_{i,j}(\lambda)) - 
\val(\chi'_M(\lambda))\big)$$
where $C_{i,j}$ denotes as above the $(i,j)$ entry of $\com(X{-}M)$.
Writing
$\com(X{-}M) = \alpha \cdot P V^t \cdot V Q^t \text{ mod } \chi_M$
as in Proposition~\ref{prop:shortcom}, we find:
\begin{align*}
N' & = \val(\alpha(\lambda)) - \val(\chi'_M(\lambda)) \\
& \hspace{2mm} + \min_{1 \leq i, j\leq n} \big(N_{j,i} + 
\val(P_i V(\lambda)^t) + \val(V(\lambda) Q_j^t)\big)
\end{align*}
where $P_i$ denotes the $i$-th row of $P$ and, similarly, $Q_j$
denotes the $j$-th row of $Q$. Note moreover that $V(\lambda)$ is
the row vector $(1, \lambda, \ldots, \lambda^{n-1})$.
By the discussion of \S \ref{}, the exact value of $M$ can be 
determined for a cost of $\softO(n^\omega)$ operations in $K$ and
$O(n^2)$ operations on integers.

When $M$ is given at flat precision, \emph{i.e.} the $N_{i,j}$'s are all 
equal to some $N$, the formula for $N'$ may be rewritten:
\begin{align*}
N' & = N + \val(\alpha(\lambda)) - \val(\chi'_M(\lambda)) \\
& \hspace{2mm} + \min_{1 \leq i \leq n} \val(P_i V(\lambda)^t)
+ \min_{1 \leq j \leq n} \val(V(\lambda) Q_j^t)\big)
\end{align*}
and can therefore now be evaluated for a cost of $\softO(n^\omega)$
operations in $K$ and only $O(n)$ operations on integers.
